---
title: "Health and Economic Impact of Weather Events in the US"
---

### Synopsis
Based on the weather events data that was provided, top health impacting weather events and the top economy impacting weather events have been determined. Two graphs have been plotted to provide a better picture. It was found that TORNADO had the biggest health impact while FLOOD had the biggest economic impact.

### Importing libraries
```{r}
library(ggplot2)
library(dplyr)
library(lubridate)
```
### Data Processing
If "storm" doesn't exist, then we read the csv file for our data
```{r}
if(!exists("storm")){
  storm <- read.csv(bzfile("repdata_data_StormData.csv.bz2"),header = TRUE)
}
```
"storm" has 902297 rows and 37 columns. Since data recorded in the earlier years is not reliable, we will use data from only the recent years for our analysis, i.e., 1990 to 2011

Introducing a "Year" column so that the new subset can be created
```{r}
storm$BGN_DATE <- as.Date(storm$BGN_DATE, "%m/%d/%Y")
storm$Year <- year(storm$BGN_DATE)
```
Creating our new 1990-2011 storm data subset
```{r}
newStorm <- storm[storm$Year>=1990,]
```
So now we have 751740 rows and 38 columns of new data

Since the first question is to find the events that are most harmful to population health, we will check for the most number of "fatalities" and "injuries". But since we are looking at their impact together, we can group "fatalities" and "injuries" into one term "pophealth"
```{r}
newStorm <- mutate(newStorm,POPHEALTH = FATALITIES+INJURIES)
newStorm$EVTYPE <- toupper(newStorm$EVTYPE)
```
Now we look at events that were the most harmful to population health
```{r}
healthImpact <- with(newStorm,aggregate(POPHEALTH~EVTYPE,FUN=sum))
subset(healthImpact,POPHEALTH>quantile(POPHEALTH,prob=0.95))
healthImpact <- newStorm %>% 
  group_by(EVTYPE) %>% 
  summarise(POPHEALTH = sum(POPHEALTH)) %>% 
  arrange(desc(POPHEALTH))
```
Therefore now we have healthImpact that contains all the EVTYPEs listed in decreasing order based on their total population health impact. 
head(healthImpact) would give the following:

EVTYPE         POPHEALTH
1 TORNADO            28426
2 EXCESSIVE HEAT      8428
3 FLOOD               7259
4 LIGHTNING           6046
5 TSTM WIND           5349
6 HEAT                3037

So we see that the most harmful EVTYPE to population health from 1990-2011 was TORNADO followed by EXCESSIVE HEAT, FLOOD, and so on 
A plot of EVTYPEs v/s health impact of the top 15 EVTYPEs would provide a better picture
```{r}
a <- ggplot(healthImpact[1:15,], aes(x=reorder(EVTYPE, -POPHEALTH),y=POPHEALTH,color=EVTYPE)) + 
  geom_bar(stat="identity", fill="yellow") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  xlab("Event") + ylab("Number of fatalities + injuries") +
  theme(legend.position="none") +
  ggtitle("Population Health impact caused by weather events")
a
```
From the graph it is visible that TORNADO had by far the most impact, almost 3 times more than the second most harmful event

### Economic Impact
Now we move on to analyse the economic impact these weather events had in the US from 1990-2011
In addition to PROPDMG and CROPDMG, their exponents PROPDMGEXP and CROPDMGEXP have also been recorded
```{r}
newStorm$PROPDMGEXP <- toupper(newStorm$PROPDMGEXP)
newStorm$CROPDMGEXP <- toupper(newStorm$CROPDMGEXP)
newStorm$CROPDMGFACTOR[(newStorm$CROPDMGEXP == "")] <- 10^0
newStorm$CROPDMGFACTOR[(newStorm$CROPDMGEXP == "?")] <- 10^0
newStorm$CROPDMGFACTOR[(newStorm$CROPDMGEXP == "0")] <- 10^0
newStorm$CROPDMGFACTOR[(newStorm$CROPDMGEXP == "2")] <- 10^2
newStorm$CROPDMGFACTOR[(newStorm$CROPDMGEXP == "K")] <- 10^3
newStorm$CROPDMGFACTOR[(newStorm$CROPDMGEXP == "M")] <- 10^6
newStorm$CROPDMGFACTOR[(newStorm$CROPDMGEXP == "B")] <- 10^9
newStorm$PROPDMGFACTOR[(newStorm$PROPDMGEXP == "")] <- 10^0
newStorm$PROPDMGFACTOR[(newStorm$PROPDMGEXP == "-")] <- 10^0
newStorm$PROPDMGFACTOR[(newStorm$PROPDMGEXP == "?")] <- 10^0
newStorm$PROPDMGFACTOR[(newStorm$PROPDMGEXP == "+")] <- 10^0
newStorm$PROPDMGFACTOR[(newStorm$PROPDMGEXP == "0")] <- 10^0
newStorm$PROPDMGFACTOR[(newStorm$PROPDMGEXP == "1")] <- 10^1
newStorm$PROPDMGFACTOR[(newStorm$PROPDMGEXP == "2")] <- 10^2
newStorm$PROPDMGFACTOR[(newStorm$PROPDMGEXP == "3")] <- 10^3
newStorm$PROPDMGFACTOR[(newStorm$PROPDMGEXP == "4")] <- 10^4
newStorm$PROPDMGFACTOR[(newStorm$PROPDMGEXP == "5")] <- 10^5
newStorm$PROPDMGFACTOR[(newStorm$PROPDMGEXP == "6")] <- 10^6
newStorm$PROPDMGFACTOR[(newStorm$PROPDMGEXP == "7")] <- 10^7
newStorm$PROPDMGFACTOR[(newStorm$PROPDMGEXP == "8")] <- 10^8
newStorm$PROPDMGFACTOR[(newStorm$PROPDMGEXP == "H")] <- 10^2
newStorm$PROPDMGFACTOR[(newStorm$PROPDMGEXP == "K")] <- 10^3
newStorm$PROPDMGFACTOR[(newStorm$PROPDMGEXP == "M")] <- 10^6
newStorm$PROPDMGFACTOR[(newStorm$PROPDMGEXP == "B")] <- 10^9
```
Now we multiply the damages with their corresponding exponents and use a single term to capture their collective contribution
```{r}
newStorm <- mutate(newStorm, ECONOMICCOST = PROPDMG * PROPDMGFACTOR + CROPDMG * CROPDMGFACTOR)
```
Using the same process we adopted to find the most harmful events for population health impact, we shall find the most harmful events for economic impact
```{r}
ecoCost <- with(newStorm, aggregate(ECONOMICCOST ~ EVTYPE, FUN = sum))
subset(ecoCost, ECONOMICCOST > quantile(ECONOMICCOST, prob = 0.95))
```
We should arrange the EVTYPEs in the decreasing order
```{r}
ecoCost <- newStorm %>% 
  group_by(EVTYPE) %>% 
  summarise(ECONOMICCOST = sum(ECONOMICCOST)) %>% 
  arrange(desc(ECONOMICCOST))
```
head(ecoCost) gives the following
EVTYPE             ECONOMICCOST
1 FLOOD             150319678257 
2 HURRICANE/TYPHOON  71913712800 
3 STORM SURGE        43323541000 
4 TORNADO            30883688776
5 HAIL               18761221986
6 FLASH FLOOD        18243991078

To see a better picture, we shall create a plot with the top 15 most harmful EVTYPEs
```{r}
b <- ggplot(ecoCost[1:15,], aes(x=reorder(EVTYPE, -ECONOMICCOST),y=ECONOMICCOST,color=EVTYPE)) + 
  geom_bar(stat="identity", fill="pink") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  xlab("Event") + ylab("Economic cost") +
  theme(legend.position="none") +
  ggtitle("Economic cost impact caused by weather events")
b
```
From the graph it is apparent that FLOOD has the most economic impact followed by HURRICANE/TYPHOON, STORM
SURGE, and so on

### Results:
FLOOD had the biggest economic impact while TORNADO had the biggest health impact. Other events and their impact
can be gauged from the plots and they have also been mentioned above.